How does the writing system get informative? Differentiation or conservation?

### 2023-06-21 manuscript/wp4.tex
ms

\documentclass[doc,biblatex]{apa7}
\DeclareLanguageMapping{american}{american-apa}
\addbibresource{references.bib}
\captionsetup[figure]{textfont=rm,font=small}
\usepackage{amsmath}
\usepackage{fontspec}
\setmainfont{Times New Roman}

@@

Is it possible, then, that heterography may be selected for in the cultural evolution of orthography? And if so, what might be the mechanism through which such selection could operate? \textcite[pp.~325--326]{Berg:2021} sketch two models of the emergence of heterography (see Fig.~\ref{}). The first model, \textit{the differentiation model}, explains heterography through the creation of a new orthographic form to differentiate between homophones. Imagine, for example, that English had chosen to adopt the French spelling <banque> for banks of the financial kind to differentiate them from banks of the river kind, resulting in a state of heterography: /baŋk/–/baŋk/ in speech, but <bank>–<banque> in writing. The second model, \textit{the conservation model}, explains heterography as the historical residue of sound change: Two spoken forms merge and become homophonous, but their heterographic spellings are conserved in the orthography. For example, the words \textit{night} and \textit{knight}---heterographic homophones in modern English---were previously pronounced differently from each other (in Old and Middle English, /kn/ was a permissible consonant cluster, as it still is in modern Dutch and German).

byte bite

bizarre bazaar

bombe

cache

<cheque>, which the OED describes as a ``a differentiated spelling of \textit{check}.''

Tindr

@@

@@

We recruited 539 participants via the Prolific platform. Participants were paid £3.00 for participation plus additional bonuses of up to £1.36 as detailed below (median bonus: £0.xxx). Participants were required to have English as a first or second language to ensure they could understand the consent form and instructions; no other restrictions were applied. Participants' most common first languages were: xxx. 44 participants (8\%) were lost to rejections or communication-game matching failures, leaving us with 495 participants in our final dataset.

@@

Participants were taught words for 16 ``alien'' objects---four shapes that could appear in four colors (see Fig.~\ref{}). The alien words had both a spoken and a written form and were composed of a stem and a suffix. The stems, which express the shape dimension, were /buv/ <buv>, /zɛt/ <zet>, /gaf/ <gaf>, and /wɒp/ <wop>. To make the stems easy to learn, they were designed to be graphically and phonetically iconic of the shapes they represent. The suffix was always pronounced /ikəʊ/ in speech, but its spelling was free to evolve over time. Thus, the spoken form of the language consists of just four unique words---/buvikəʊ/, /zɛtikəʊ/, /gafikəʊ/, and /wɒpikəʊ/---that mark a shape distinction; however, the \textit{spelling} of the suffix could potentially take on different forms to mark color. The spoken forms were synthesized using the Apple text-to-speech synthesizer (the Tessa voice).

The transmission chains were seeded with an orthographic system that may exhibit high variation or no variation according to condition. In the no-variation conditions, the suffix had only one spelling, although each chain was initialized with a different randomly generated spelling. To generate a spelling we randomly choose a grapheme from each of the sets \{<ee>, <i>, <y>\}, \{<c>, <ck>, <k>\}, and \{<oe>, <oh>, <ow>\} and concatenate them. For example, in one chain the suffix might initially be spelled <icoe>, while in another chain it might initially be spelled <yckow>. In the high-variation conditions, the spellings were randomly generated not just for each chain but also for each of the 16 objects. Thus, although the star shaped objects are all called /zɛtikəʊ/ in speech, the different colors will all have different random spellings (e.g., <zetikow> for pink, <zeticoe> for gray, <zeteeckoe> for blue, and <zetyckoh> for yellow). These spellings were not systematic, however; the <yckoh> spelling was not systematically used to represent yellow across different shapes. Essentially, then, in the no-variation conditions there is initially only one possible way to spell /ikəʊ/, while in the high-variation conditions there are many possible ways to spell it without any systematic pattern.

@@

As outlined above, the participants were arranged into transmission chains such that the spellings produced by one participant would subsequently be taught to the next participant in the chain. The first participant in a chain was taught the randomly generated orthographic system explained in the previous section, and this system was then free to evolve as it was culturally transmitted. This was subject to a bottleneck on transmission: Not all 16 spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only 12 of the 16 spellings generated at generation $i-1$ (at least one of each shape and at least one of each color). This procedure continued until the chain converged on a particular orthographic system. A chain is said to converge when two consecutive generations produce identical spellings for all 16 items, suggesting that the chain is unlikely to experience significant further change. We adopted this convergence-based approach because we were interested in revealing what kinds of orthographic system are stable under the four different conditions.

@@

Participants were trained on the spoken and written forms through a combination of passive exposure trials and mini-test trials. In the passive exposure trials, the alien object appeared first, accompanied by the written and spoken forms after a 500~ms delay. In the mini-test trials, which were interleaved among the passive exposure trials, the participant had to type in the appropriate written form for one of the objects. The participant could freely type in a word and would then be given feedback on any spelling errors (additions shown with green text and deletions shown with red strikethrough text). Participants received a 1p bonus for spelling the stem correctly or a 2p bonus for spelling the entire word correctly.

\subsubsection{Test procedure}

\subsubsection{Communication game procedure}

@@

@@

@@

%    ORIG.    SC1      SC2      SC3

% P: /ɪθə/           > /ɪθəʊ/ > /ɪθɛɪ/
% G: /ɪsɛɪ/ > /ɪhɛɪ/          > /ɪθɛɪ/
% B: /ɪʃəʊ/          > /ɪθəʊ/ > /ɪθɛɪ/
% Y: /ɪhu/  > /ɪhɛɪ/          > /ɪθɛɪ/

>>>>

\PassOptionsToPackage{hidelinks}{hyperref}
\documentclass[doc,biblatex]{apa7}
\DeclareLanguageMapping{american}{american-apa}
\addbibresource{references.bib}
\captionsetup[figure]{textfont=rm,font=small}
\usepackage{amsmath}
\usepackage{mathspec}
\setallmainfonts{Times New Roman}
\setmainfont{Times New Roman}

@@

Is it possible, then, that heterography may be selected for in the cultural evolution of orthography? And if so, what might be the mechanism through which such selection could operate? \textcite[pp.~325--326]{Berg:2021} sketch two models of how a word might enter a state of heterography, as depicted in Fig.~\ref{berg_aronoff_models}. The first model, \textit{the differentiation model}, explains heterography through the creation of a new orthographic form. For example, in (British) English, <cheque>, which the OED describes as a ``a differentiated spelling of \textit{check},'' allows the written language to distinguish the bank order from other meanings of the word \textit{check}. The word \textit{byte} is a respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}. In English, surnames that are homophonous with ordinary nouns often add <e> (Clarke, Greene, Wilde) or double the final consonant (Carr, Hogg, Mann), and it is common for brand names to adopt creative spellings, such as <flickr> (from <flicker>), <Froot Loops> (from <fruit>), or MVMT (from <movement>).

The second model, \textit{the conservation model}, explains heterography as the historical residue of sound change: Two spoken forms merge and become homophonous, but their heterographic spellings are conserved in the orthography. For example, the words \textit{night} and \textit{knight}---heterographic homophones in modern English---were previously pronounced differently from each other (/niçt/ and /kniçt/; in Old and Middle English, /kn/ was a permissible consonant cluster, as it still is in modern Dutch and German).

@@

\subsection{Research questions}

Q1) Does a learning pressure (induced through cultural transmission) result in the emergence of a simple, systematic, transparent orthography that is easier to learn? Hypotheses: In the high-variation conditions, where the orthography starts out chaotic (resulting in holistic systems), we expect simpler systems to emerge. Simplicity may be realized in a number of ways. Transparent systems use a one-to-one mapping to directly represent the spoken forms in writing. Redundant systems retain some spelling variation (with idiosyncratic spellings of the suffix for each shape) but become simpler by collapsing across colour. Expressive systems continue to express colour, but do so with a non-transparent but easy to learn compositional suffix system. The prevalence of these three types of system will be influenced by the presence of the communicative pressure. When communication is not present, transparent systems will tend to dominate, alongside some redundant systems and possibly some expressive systems. When communication is present, expressive systems will tend to dominate, perhaps with some redundant and holistic forms. The gradual emergence of simplicity, whatever form it takes, should be accompanied by increased learnability.

Q2) Can a writing system evolve to become more informative than its uninformative spoken counterpart? Hypotheses: Yes, but preferentially when two vital ingredients are present: a communicative pressure for informativeness (which makes an expressive system desirable) and high variation (which provides the building blocks to make that happen). When both ingredients are present, expressive systems will tend to dominate. If only one of the ingredients is available (either communication or high variation), informative systems may emerge but are less likely. In the case of learning-only and high-variation, expressive systems may emerge if the learners – yearning for an explanation for the variation – assume that colour must somehow be involved and therefore impose (perhaps unconsciously) colour-expressive structure on the language (i.e. learners may, to a certain extent, have a preconceived notion that languages ought to be informative). In the case of communication and no variation (where, again, only one of the ingredients is met), expressive systems may emerge but this is highly dependent on participants being 'innovators' who are willing to develop novel spelling systems that diverge from the spoken system in order to express colour. If neither ingredient is present, expressive systems will not emerge.

@@

We recruited 538 participants via the Prolific platform. Participants were paid £3.00 for participation plus additional bonuses of up to £1.36 as detailed below (median bonus: £0.64). The median completion time was 19:36, resulting in an hourly rate of £9.18 (£11.03 including bonus). We limited recruitment to participants who had English as a first or second language to ensure they could understand the consent form and instructions; no other restrictions were applied. The most common first languages were: English (59\%), Polish (9.7\%), Portuguese (6.1\%), Italian (5.4\%), Spanish (5\%), and Greek (3.7\%). Five participants were rejected because they did not pass the random audio checks, and a further three participants were removed because they had been partnered with one of the five rejected participants. In addition, 35 participants were lost to communication-game pairing failures, resulting in a final dataset of 495 participants.

@@

Participants were taught words for 16 ``alien'' objects---four shapes that could appear in four colors (see Fig.~\ref{}). The alien words had a spoken and written form composed of a stem and suffix. The stems, which express the shape dimension, were /buv/ <buv>, /zɛt/ <zet>, /gaf/ <gaf>, and /wɒp/ <wop>. Since we are not interested in participants' learning of the stems, they were designed to be quick and easy to learn by being graphically and phonetically iconic of the shapes they represent, and they were fixed and unchanging throughout the experiment. The suffix was always pronounced /ikəʊ/ in speech, but its spelling was free to change over time. Thus, the spoken form of the language consists of just four unique words---/buvikəʊ/, /zɛtikəʊ/, /gafikəʊ/, and /wɒpikəʊ/---that mark a shape distinction; however, the \textit{spelling} of the suffix could potentially take on different forms to mark color. The spoken forms were synthesized using the Apple text-to-speech synthesizer (the Tessa voice).

The transmission chains were seeded with an orthographic system that may exhibit high variation or no variation according to condition. In the no-variation conditions, the suffix had only one spelling, although each chain was initialized with a different randomly generated spelling. To generate a spelling we randomly choose a grapheme from each of the sets \{<ee>, <i>, <y>\}, \{<c>, <ck>, <k>\}, and \{<oe>, <oh>, <ow>\} and concatenate them. For example, in one chain the suffix might initially be spelled <icoe>, while in another chain it might initially be spelled <yckow>. In the high-variation conditions, the spellings were randomly generated not just for each chain but also for each of the 16 objects. Thus, although the star shaped objects are all called /zɛtikəʊ/ in speech, the different colors have different random spellings (e.g., <zetikow> for pink, <zeticoe> for gray, <zeteeckoe> for blue, and <zetyckoh> for yellow). These spellings were not systematic, however; for example, the <yckoh> spelling was not systematically used to represent yellow across different shapes. Essentially, then, in the no-variation conditions there is initially only one possible way to spell /ikəʊ/, while in the high-variation conditions there are many possible ways to spell it without any systematic pattern.

@@

As outlined above, the participants were arranged into transmission chains such that the spellings produced by one participant would subsequently be taught to the next participant in the chain. The first participant in a chain was taught the initial, randomly generated orthographic system, and this system was then free to evolve as it was culturally transmitted. This was subject to a bottleneck on transmission: Not all 16 spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only 12 of the 16 spellings generated at generation $i-1$ (at least one of each shape and at least one of each color). Transmission continued until the chain converged on a particular orthographic system. A chain is said to converge when two consecutive generations produce identical spellings for all 16 items (including the four unseen items), suggesting that the chain is unlikely to experience significant further change. We adopted this convergence-based approach because we were primarily interested in revealing what kinds of orthographic system are stable under the four conditions, but it was unclear a~priori how many generations would be required to reveal this information.

@@

Participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials. In the passive exposure trials, the alien object was presented alongside the written and spoken forms for 2000~ms. The written form appeared after a 500~ms delay so that participants would first attend to the object before looking at the word. In mini-test trials, which were interleaved among the passive exposure trials, the participant had to type in the appropriate written form for one of the objects. The mini-tests were included to help participants actively engage in the learning process. After submitting a spelling, the participant was given feedback on any errors (additions shown in bold green text and deletions shown in red strikethrough text). The input box only accepted alphabetical strings of between four and nine characters, and spellings containing certain letter combinations (e.g., <bk>, <bl>, <pin>, <ye>) were not accepted to prevent participants from entering English color words or abbreviations thereof. Participants received a 1p bonus for spelling the stem correctly or a 2p bonus for spelling the entire word correctly. However, to receive the bonus, the word had to be submitted within 20~seconds. Each of the 12 object--word pairs in the training set (i.e., those that passed through the bottleneck on transmission) was passively exposed 12 times and mini-tested 3 times, resulting in a total of 144 passive exposure trials and 36 mini-test trials (the maximum possible bonus in training was therefore 72p). To ensure that participants were indeed listening to the spoken forms, a voice asked participants to click on the object at three random points during training; participants who did not follow this instruction were excluded.

\subsubsection{Non-communicative test procedure}

After training, participants assigned to the non-communicative conditions completed a test phase, alternating between two types of test trial, production trials and comprehension trials. In production trials, the participant was shown an object and heard its associated word pronounced aloud. The task was to type~in the appropriate spelling under the same input restrictions as in mini-test trials and an additional restriction that the stem (i.e., the first three letters of the word) had to be spelled correctly to ensure that stems could not diverge from the spoken forms over time. Since participants heard the word pronounced aloud, typing the stem correctly should be trivial, but in cases where the stem was initially spelled incorrectly, a popup message explicitly told the participant the correct spelling of the stem. In comprehension trials, the participant was shown a word and had to click on the matching object from all 16 objects arranged in random order (in cases where multiple objects were described by the same wordform, any of the objects was a valid choice). In both types of test trial, the participant was awarded a bonus of 2p for every correct answer (unlike mini-tests, there was no 1p bonus for partially correct spellings). Each of the 16 object--word pairs (i.e., including the four unseen items) was tested once in production and once in comprehension, resulting in 32 trials and a maximum bonus of 64p. Binary feedback was provided on each trial in the form of a sound effect.

\subsubsection{Communicative test procedure}

After training, participants assigned to the communicative conditions completed a live, over-the-internet communication game with another participant. The two participants both received training on the same orthographic system inherited from a previous generation. The communication game mirrors the overall structure of the non-communicative test with the production and comprehension trials becoming the two sides of a communicative interaction. On a given trial, one participant---the sender---would complete a production trial (produce a word for a given meaning) and this word was transmitted to the other participant. The second participant---the receiver---would then complete a comprehension trial (select an object for the transmitted word). The participants then switched roles, resulting in the same overall trial structure as the non-communicative condition (i.e., alternation between production and comprehension trials with the same bonusing scheme). However, the communication game differed from the non-communicative test in three ways. First, the transmitted and received words were displayed in text message speech bubbles to emphasize the communicative nature of the test. Second, in cases where a single wordform maps to multiple objects, the receiver performing the comprehension trial could not click any of the objects---their goal was to click the target object being communicated by the sender. Third, both participants received rich feedback: The sender saw which object the receiver clicked on, and the receiver saw which object was the correct target. Overall, then, the communicative test is identical to the non-communicative test in terms of the task to be performed, but the goal is quite different: In the non-communicative test, the goal and reward structure is based on accurately reproducing the orthography learned during training, whereas in the communicative test, the goal and reward structure is based on successfully communicating a target item.

@@

For the purpose of the results, we are primarily interested in the spelling of the suffixes, so we first needed to isolate the suffixes from the stems. Although the first three letters (<buv>, <zet>, <gaf>, <wop>) were fixed and unchanging throughout the experiment, it was possible for the stems to grow longer if additional material was inserted before the suffix (e.g., <gaff>, <wopb>, <zetch>). We consider the stem to be all material up to but not including the first vowel letter after the original three-letter stems, such that, for example, <zethyckoh> will be parsed into the stem <zeth> and suffix <yckoh>. Of course, it is not possible to know how participants themselves divided the strings (if at all), but since all words were phonetically identical from the /i/ vowel onward, it is reasonable to parse the strings this way on the basis that this yields the shortest grammar.

To analyze the variation in the suffix spellings and how it correlates with meaning, each participant's system was transformed into a 4×4 matrix, with rows representing the shape dimension, columns representing the color dimension, and the matrix values representing the unique spellings. For example, a ``degenerate'' system of suffix spellings---one that uses a single spelling for all meanings---would have the following matrix representation:
\begin{equation}
	S_\mathrm{degenerate} = \begin{bmatrix}
	1 & 1 & 1 & 1 \\
	1 & 1 & 1 & 1 \\
	1 & 1 & 1 & 1 \\
	1 & 1 & 1 & 1
	\end{bmatrix}
\end{equation}
Here the number 1 represents the one and only suffix spelling. We define an ``expressive'' system as one that marks the color dimension using four distinct spellings, resulting in the following matrix:
\begin{equation}
	S_\mathrm{expressive} = \begin{bmatrix}
	1 & 2 & 3 & 4 \\
	1 & 2 & 3 & 4 \\
	1 & 2 & 3 & 4 \\
	1 & 2 & 3 & 4
	\end{bmatrix}
\end{equation}
In such a system, although the suffix only expresses color, the words as a whole are fully informative, since shape is expressed by the stem. A third system of interest is the ``redundant'' system:
\begin{equation}
	S_\mathrm{redundant} = \begin{bmatrix}
	1 & 1 & 1 & 1 \\
	2 & 2 & 2 & 2 \\
	3 & 3 & 3 & 3 \\
	4 & 4 & 4 & 4
	\end{bmatrix}
\end{equation}
Although $S_\mathrm{redundant}$ carries the same amount of information as $S_\mathrm{expressive}$, it marks a dimension (shape) that is already expressed by the stem, making the information carried by the suffix redundant. Such a system contains unnecessary complexity---different ways of writing /ikəʊ/ that carry no additional information.

Of course, the suffix spelling systems produced by participants may not match these three reference systems exactly. To determine which reference system a participant's output system is most similar to, we adopt variation of information \parencite[VoI;][]{Meila:2007} as a distance metric. VoI is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one set partition into another set partition. So, for example, the distance between $S_\mathrm{degenerate}$ and $S_\mathrm{expressive}$ is 2~bits because $S_\mathrm{expressive}$ conveys 2~bits of information\footnote{In $S_\mathrm{expressive}$, knowing the suffix reduces uncertainty from 16 equally probable meanings to 4 equally probable meanings; $-\log\frac{4}{16} = 2$~bits.}, 2~bits more than $S_\mathrm{degenerate}$, which conveys 0~bits of information. The distance between $S_\mathrm{expressive}$ and $S_\mathrm{redundant}$ is 4~bits; this is because, although both systems convey 2~bits of information, the information they convey is entirely orthogonal; in switching from $S_\mathrm{expressive}$ to $S_\mathrm{redundant}$, one loses 2~bits of information (information about color), but one gains 2~bits of information (information about shape), making the total distance 4~bits.

We computed the VoI between the emergent suffix systems and the three reference systems and plotted them on a ternary diagram, the three vertices of which represent the three reference systems (see Fig.~\ref{exp1_ternary_plot}). The diagram is divided into three sectors (shades of gray), which reveal which reference system the final converged-on suffix systems are most similar to. Most of the emergent system are either redundant or degenerate, with some straddling the border between the two. No system is classified as expressive, although two of the converged-on systems (highlighted in Fig.~\ref{exp1_ternary_plot}) showed some signs of color-marking. In one system, <icoe> is used for gray objects and <ikoe> for all other colors. In the other system, <ikcow> is used for gray objects and <ikoe> is used for all other colors, except the star shape uses <y> instead of <i> (i.e., <ykcow> and <ykoe>), resulting in some spelling redundancy\footnote{However, rather than parsing the stems as <buv>, <zett>, <gaf>, <wop>, they could instead be parsed as <buvi>, <zetty>, <gafi>, <wopi>. Under this interpretation, there would be no redundancy and the system would be equivalent to the other 1-color emergent system.} In general, there is no discernible difference between the two conditions; regardless of whether the communicative pressure was present, the emergent suffix spelling systems tended to fall along the degenerate--redundant axis, and the only two cases of color-marking were in~fact from the learning-only condition.

@@

Our results suggest that the differentiation model is not a good explanation for the emergence of informative heterography.

Three pathways

- conscious design by communicators
- emergence through learning (chance alignment with color)
-

@@
